### 2015-05-08 papers/histogram/thesis/thesis.tex
added back wang landau, and finished first complete draft of thesis

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Previous
works have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We propose to implement these
methods in Monte Carlo simulations of the square-well fluid, determine
appropriate performance metrics for their comparison, and identify
those methods which are most efficient and effective at sampling
otherwise improbable system states.\fixme{make this an abstract,
  rather than a proposal...}

\fixme{Rewrite after the rest of the thesis is completed}

\fixme{Always talk about what we do in our own code in present, rather
  than past tense}

@@

In our implementation of the simple flat method, we used $u=2$ and
$n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is the number of
energy levels of the given square-well fluid and $T_{\t{min}}$ is a
minimum temperature of interest for the current simulation. The factor
of $L$ appears because initialization time should scale with the range
of energies of the simulated system. If we have twice the number of
energies, we should explore energy space for twice as long. To
estimate $L$, we multiplied the maximum number of spheres $M$ which
fit within a radius of $\lambda\sigma$ of a single sphere on a
face-centered cubic lattice (i.e. the maximum number of spheres with
which a given sphere could interact) by the total number of spheres
$N$, and divided the result by two so as to not double-count
interactions between spheres (so $L=MN/2$). The exponential
$\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for two adjacent
energies, appears because lower minimum temperatures of interest
warrant sampling lower energies, which in turn requires simulating for
longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (discussed in Section \ref{sec:min_energy}). In
this paper, we will always use $kT_{\t{min}}=0.2\epsilon$.

@@

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we effectively
forbid transitions to energies $E<E_-$ by tentatively setting weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT}$ for some temperature $T\ll
T_{\t{min}}$. These weights prevent the initialization routine from
wasting time on energies at which we will override the weight array
after initialization anyways. A special case was made for Wang-Landau,
as the design of this method assumes knowledge of the energy range of
interest a priori. In order to fairly compare Wang-Landau to other
methods, we therefore identified $E_-$ for each simulated system using
other methods, and hard-coded the minimum important energies for each
system into the Wang-Landau initialization routines\fixme{talk about
  how bad this is}. We always set
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ after initialization of all
histogram methods with $kT_{\t{min}}=0.2\epsilon$. We chose this
temperature because it is right below the temperature at which the
square-well fluid condenses from a gas into a liquid (see Section
\ref{sec:errors}). \fixme{what about pressure?}

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). To increase confidence in
low-temperature gold standard results, we initialized them with
$kT_{\t{min}}=0.1\epsilon$. After initialization, the gold standard
runs simulate for $10^{10}$ iterations ($10^{10}N$ moves). While the
entire initialization and simulation process takes around two hours
for the largest regular simulations, the largest golden simulations
initialize for $\sim11$ hours, and simulate for a week. To control for
iteration number in the quality of results, we compare against a
``converged flat'' simulation. Converged flat simulations use the
golden (flat histogram) weight array, but simulates for the same
number of iterations as the standard simulations.

@@

\begin{figure}[!b]
  \centering
  \includegraphics[width=\textwidth]{figs/array-example.pdf}
  \caption[Example arrays]
  {Example arrays computed by various histogram methods for a
    square-well fluid with a well width $\lambda=1.3$, filling
    fraction $\eta=0.3$, and $N=25$ spheres after simulating for
    $10^5N^2$ iterations: (top left) energy histogram, (top right)
    weight array, (bottom left) density of states. Vertical reference
    lines denote the minimum important energy $E_-$.}
  \label{fig:arrays}
\end{figure}

Most of our analysis will use quantities computed from two arrays
given by simulations: the energy histogram $H\p{E}$ and the weights
$w\p{E}$. To get a sense of what these arrays look like, Figure
\ref{fig:arrays} shows the energy histogram and weights from an $N=25$
simulation, along with the density of states computed via $\tilde
D=H/w$ and normalized to have a maximum value of 1. The vertical
reference line in these plots denotes the minimum important energy
$E_-$. These plots demonstrate that, as expected, different methods
can accumulate different histograms $H$ and weights $w$, yet agree on
system properties such as the density of states $D$.

Though high energy histogram counts at low energies mean that a
simulation has collected many statistics on these energies, the energy
histogram says nothing about how correlated different energy
observations are. To determine how effective different simulations are
at collecting uncorrelated statistics, we can look at the pessimistic
sample rate $r_p\p{E}$, which is the mean number of simulation
iterations required to sample a given energy $E$. This rate is given
in terms of the number of iterations $I$ in a simulation (where we
always have $I=10^5N^2$) and the pessimistic sample rate $s_p\p{E}$ by
$r_p\p{E}=I/s_p\p{E}$. Figure \ref{fig:sample_rates} shows pessimistic
sample rates for all methods averaged over 30 random number seeds for
$N=25$ simulations. Lower pessimistic sample rates mean that a
simulation takes less time to collect statistics which are guaranteed
to be uncorrelated.

\fixme{talk about plots}

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.8\textwidth]{figs/sample-rates.pdf}
  \caption[Example energy sampling rates]
  {Example pessimistic energy sampling rates computed by various
    histogram methods. Sample rates for each method are averaged over
    30 simulations with different random number seeds. The vertical
    reference line denotes the minimum important energy $E_-$.}
  \label{fig:sample_rates}
\end{figure}

@@

We can use the formula in (\ref{eq:XT}) to compute the temperature
dependence of a few system properties directly from the density of
states, such as the internal energy
\begin{align}
  U=\bk{E}_T,
  \label{eq:U}
\end{align}
the heat capacity
\begin{align}
  C_V\p{T}=\f{\partial U}{\partial T}
  =\bk{\p{\f{E}{kT}}^2}_T-\bk{\f{E}{kT}}_T^2,
  \label{eq:CV}
\end{align}
and the configurational entropy\fixme{how is this different from
  regular entropy?}
\begin{align}
  S_{\t{config}}\p{T}=k\bk{\f{E}{kT}+\ln Z}_T.
  \label{eq:S}
\end{align}
Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} show these
properties as computed via all methods for a single $N=25$ simulation
each. When looking at these figures, one should remember that results
are only trusted down to $kT=0.2\epsilon$ for regular simulations, and
$kT=0.1\epsilon$ for golden ones. As with the density of states, all
histogram methods agree fairly well on the physical properties of the
square-well fluid. Though Figures
\ref{fig:internal_energy}--\ref{fig:config_entropy} are interesting
for studying the square-well fluid, they do not help us resolve the
efficacy of different histogram methods. These figures do, however,
show that the square-well fluid has a phase transition around
$kT\approx0.5\epsilon$, which justifies our choice of
$kT_{\t{min}}=0.2\epsilon$ and ensures that the histogram methods are
performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to demonstrate that this phase transition is
indeed between liquid and gas, and take this fact as given.

Having a sense of how $U$, $C_V$ and $S_{\t{config}}$ depend on
temperature, we can now look at the how the errors $\Delta U$, $\Delta
C_V$ and $\Delta S_{\t{config}}$ relative to the TMMC golden
calculations. These are shown for a single $N=25$ simulation in
Figures \ref{fig:internal_energy_err}--\ref{fig:config_entropy_err}.

\begin{figure}[b]
  \centering
  \includegraphics[width=0.7\textwidth]{figs/internal-energy.pdf}
  \caption[Internal energy]
  {Specific internal energy curves from single $N=25$ simulations.}
  \label{fig:internal_energy}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/heat-capacity.pdf}
  \caption[Heat capacity]
  {Specific heat capacity curves from single $N=25$ simulations.}
  \label{fig:heat_capacity}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/config-entropy.pdf}
  \caption[Configurational entropy]
  {Specific configurational entropy curves from single $N=25$
    simulations.}
  \label{fig:config_entropy}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]
  {figs/internal-energy-err.pdf}
  \caption[Internal energy error scaling]
  {Errors in specific internal energy from single $N=25$ simulations.}
  \label{fig:internal_energy_err}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/heat-capacity-err.pdf}
  \caption[Heat capacity error scaling]
  {Errors in specific heat capacity from single $N=25$ simulations.}
  \label{fig:heat_capacity_err}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/config-entropy-err.pdf}
  \caption[Configurational entropy error scaling]
  {Errors in specific configurational entropy from single $N=25$
    simulations.}
  \label{fig:config_entropy_err}
\end{figure}

\subsubsection{Error scaling}
\label{sec:scaling}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/u-scaling.pdf}
  \caption[Internal energy error scaling]
  {Errors in internal energy as computed by various histogram methods,
    each averaged over 30 simulations with different random number
    seeds.}
  \label{fig:u_scaling}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/cv-scaling.pdf}
  \caption[Heat capacity error scaling]
  {Errors in specific heat capacity as computed by various histogram
    methods, each averaged over 30 simulations with different random
    number seeds.}
  \label{fig:cv_scaling}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/s-scaling.pdf}
  \caption[Configurational entropy error scaling]
  {Errors in specific configurational entropy as computed by various
    histogram methods, each averaged over 30 simulations with
    different random number seeds.}
  \label{fig:s_scaling}
\end{figure}

\subsubsection{Error comparisons}
\label{sec:errors}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]
  {figs/internal-energy-comps.pdf}
  \caption[Internal energy error comparisons]
  {Internal energy error comparisons}
  \label{fig:u_comps}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/heat-capacity-comps.pdf}
  \caption[Heat capacity error comparisons]
  {Heat capacity error comparisons}
  \label{fig:u_comps}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/config-entropy-comps.pdf}
  \caption[Configurational entropy error comparisons]
  {Configurational entropy error comparisons}
  \label{fig:u_comps}
\end{figure}

@@

\fixme{What have we learned? What recommendations can we make? When is
  it appropriate to use which histogram methods?}

>>>>

We identify and develop efficient numerical methods for determining
thermodynamic properties of the square-well fluid in order to test
square-well density functional theories. The liquid-vapor phase
interface is an interesting regime for testing density functional
theories, but unbiased Metropolis Monte Carlo simulations are
incapable of sampling the low energy fluid states which dominate the
partition function at low (i.e. liquid-state) temperatures. Previous
works have developed several generic Monte Carlo histogram methods for
collecting statistics on low energy system states, but little or no
literature exists on these methods' systematic comparison, as well as
their application to the square-well fluid. The square-well fluid in
particular introduces application challenges not manifest in
traditional models for testing and benchmarking such numerical
techniques (e.g. the Ising model). We implement the Wang-Landau
method, transition matrix Monte Carlo, as well as two methods of our
own. The first of our methods is novel, while the second is a hybrid
of two published methods (transition matrix Monte Carlo and the
optimized ensemble).

We find that Wang-Landau is capable of performing the best of all
methods, but fails catastrophically without a predefined energy range
for the square-well fluid. Our first method gives results comparable
to successful Wang-Landau simulations, but with some anomalous
cases. Our implementation of the optimized ensemble using a transition
matrix results in worse performance over the straightforward
transition matrix Monte Carlo method.

\fixme{edit when I'm awake}

@@

In our implementation of the simple flat method, we used $u=2$ and
$n_0=L\exp\p{\epsilon/kT_{\t{min}}}$, where $L$ is the number of
energy levels of the given square-well fluid and $T_{\t{min}}$ is a
minimum temperature of interest for the current simulation. The factor
of $L$ appears because initialization time should scale with the range
of energies of the simulated system. If we have twice the number of
energies, we should explore energy space for twice as long. To
estimate $L$, we multiplied the maximum number of spheres $M$ which
fit within a radius of $\lambda\sigma$ of a single sphere on a
face-centered cubic lattice (i.e. the maximum number of spheres with
which a given sphere could interact) by the total number of spheres
$N$, and divided the result by two so as to not double-count
interactions between spheres (so $L=MN/2$). The exponential
$\exp\p{\epsilon/kT_{\t{min}}}$, a Boltzmann factor for two adjacent
energies, appears because lower minimum temperatures of interest
warrant sampling lower energies, which in turn requires simulating for
longer (in step \ref{alg:simple_flat_sim} of Algorithm
\ref{alg:simple_flat}) because these energies have lower state
densities. The introduction of a minimum temperature may seem ad-hoc,
but all simulations in this paper require a choice of minimum
temperature anyways (discussed in Section \ref{sec:min_energy}).

@@

Once we identify an energy $E_-$ satisfying (\ref{eq:min_E_condition})
in an initialization routine for any histogram method, we effectively
forbid transitions to energies $E<E_-$ by tentatively setting weights
$w\p{E<E_-}=e^{-\p{E-E_-}/kT}$ for some temperature $T\ll
T_{\t{min}}$. These weights prevent the initialization routine from
wasting time on energies at which we will override the weight array
after initialization anyways. We always set
$w\p{E<E_-}=e^{-\p{E-E_-}/kT_{\t{min}}}$ after initialization of all
histogram methods with $kT_{\t{min}}/\epsilon=0.2$. We chose this
temperature because it is right below the temperature at which the
square-well fluid condenses from a gas into a liquid (see Section
\ref{sec:properties}). \fixme{what about pressure?}

We made a special case for Wang-Landau, as the design of this method
assumes knowledge of the energy range of interest a priori. In order
to more fairly compare Wang-Landau to other methods, we therefore
simulated two cases: one in which the minimum important energy $E_-$
is identified via other methods and hard-coded into the Wang-Landau
initialization routine, and one in which Wang-Landau simply assumes
that the lowest energy a given simulation has ever observed is the
minimum important energy. The simulations in which $E_-$ is hard-coded
we will refer to as ``vanilla Wang-Landau'' simulations.

@@

Our analysis will involve comparing errors in system properties
computed from simulation results. As we do not know the true values of
these system properties, we will compare standard simulation results
against those from ``gold standard'' simulations, which run for
significantly longer than the standard $10^5N^2$ iterations. For our
gold standard simulations, we use the TMMC method and initialize until
it collects $10^4$ optimistic samples of the minimum important energy
(i.e. until $s_o\p{E_-}\ge 10^4$). \fixme{why TMMC?} To increase
confidence in low-temperature gold standard results, we initialized
them with $kT_{\t{min}}/\epsilon=0.1$. After initialization, the gold
standard runs simulate for $10^{10}$ iterations ($10^{10}N$
moves). While the entire initialization and simulation process takes
around two hours for the largest regular simulations, the largest
golden simulations initialize for $\sim11$ hours, and simulate for a
week. To control for iteration number in the quality of results, we
compare against ``converged flat'' simulations. Converged flat
simulations read in the transition matrix from golden simulations
(which accumulates statistics in both during and after initialization)
to construct a weight array, but simulate for the same number of
iterations as the standard simulations.

@@

Most of our analysis will use quantities computed from two arrays
given by simulations: the energy histogram $H\p{E}$ and the weights
$w\p{E}$. To get a sense of what these arrays look like, Figure
\ref{fig:arrays} shows the energy histogram and weights from an $N=25$
simulation, along with the density of states computed via $\tilde
D=H/w$ and normalized to have a maximum value of 1. The vertical
reference line in these plots denotes the minimum important energy
$E_-$. For reference, we also provide the histogram produced by an
infinite temperature (no weights) simulation.

The most immediate observation in these plots is the failure of
Wang-Landau to produce meaningful results without a predefined energy
range. This failure occurs because the Wang-Landau algorithm does not
have a mechanism for properly setting and adjusting weights on newly
observed energies. After several reset cycles of the Wang-Landau
algorithm, the weight-modifying factor $f_k$ gets exceedingly small,
causing Wang-Landau to get stuck at newly encountered low energies
trying to fix their weights with a factor $f_k$ whose magnitude is
appropriate for fine tuning, rather than heavy adjustment. If
Wang-Landau knows the energy range of interest to begin with, it will
not decrease $f_k$ until it has observed all important energies, and
adjusted their weights to some reasonable value. Due to the drastic
failure of Wang-Landau, will omit its results, which are nonsensical,
from many figures.

The energy histogram in Figure \ref{fig:arrays} shows that TMMC and
OE-TMMC spend considerably less time at important low energies than
the simple flat and vanilla Wang-Landau methods. At face value, this
behavior of TMMC and OE-TMMC is undesirable.

High energy histogram counts at low energies mean that a simulation
has collected many statistics on these energies, but the energy
histogram contains no information about how correlated different
energy observations are. To determine how effective different
simulations are at collecting uncorrelated statistics, we can look at
the pessimistic sample rate $r_p\p{E}$, which is the mean number of
simulation iterations required to sample a given energy $E$. This rate
is given in terms of the number of iterations $I$ in a simulation
(where we always have $I=10^5N^2$) and the pessimistic sample rate
$s_p\p{E}$ by $r_p\p{E}=I/s_p\p{E}$. Figure \ref{fig:sample_rates}
shows pessimistic sample rates for all methods except Wang-Landau
averaged over 30 random number generator seeds for $N=25$
simulations. Lower pessimistic sample rates mean that a simulation
takes less time to collect statistics which are guaranteed to be
uncorrelated. By this metric, TMMC and OE-TMMC again perform worse
than simple flat and vanilla Wang-Landau.\fixme{comment on OETMMC
  doing worse than TMMC}

\begin{figure}
  \centering
  \includegraphics[width=\textwidth]{figs/array-example.pdf}
  \caption[Example arrays]
  {Example arrays computed by various histogram methods for a
    square-well fluid with a well width $\lambda=1.3$, filling
    fraction $\eta=0.3$, and $N=25$ spheres after simulating for
    $10^5N^2$ iterations: (top left) energy histogram, (top right)
    weight array, (bottom left) density of states. Vertical reference
    lines denote the minimum important energy $E_-$.}
  \label{fig:arrays}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth]{figs/sample-rates.pdf}
  \caption[Example energy sampling rates]
  {Example pessimistic energy sampling rates computed by various
    histogram methods. Sample rates for each method are averaged over
    30 simulations with different random number seeds. The vertical
    reference line denotes the minimum important energy $E_-$.}
  \label{fig:sample_rates}
\end{figure}

@@

Histograms and sampling rates let us speculate about the quality of a
simulations, but the most important task for any simulation is to
yield accurate physical results. We therefore compute, via the formula
in (\ref{eq:XT}), the temperature dependence of the square-well
fluid's potential energy
\begin{align}
  U=\bk{E}_T,
  \label{eq:U}
\end{align}
the heat capacity
\begin{align}
  C_V\p{T}=\f{\partial U}{\partial T}
  =\bk{\p{\f{E}{kT}}^2}_T-\bk{\f{E}{kT}}_T^2,
  \label{eq:CV}
\end{align}
and the configurational entropy\fixme{how is this different from
  regular entropy?}
\begin{align}
  S_{\t{config}}\p{T}=k\bk{\f{E}{kT}+\ln Z}_T
  -k\bk{\f{E}{kT}+\ln Z}_{T\to\infty}.
  \label{eq:S}
\end{align}
All of these properties are directly computable from the density of
states $D\p{E}$, and are shown in Figures
\ref{fig:internal_energy}--\ref{fig:config_entropy} for $N=25$
simulations, averaged over 30 random number generator seeds. When
looking at these figures, one should remember that we are only
confident in results down to $kT/\epsilon=0.2$ for regular
simulations, and $kT/\epsilon=0.1$ for golden ones. With the exception
of Wang-Landau, which has a heavy low-energy bias, all simulations
agree fairly well on these properties of the square-well fluid.

Though Figures \ref{fig:internal_energy}--\ref{fig:config_entropy} are
interesting for studying the square-well fluid, they do not help us
resolve the efficacy of different histogram methods. These figures do,
however, show that the square-well fluid has a phase transition around
$kT/\epsilon\approx0.5$, which justifies our choice of
$kT_{\t{min}}/\epsilon=0.2$ and ensures that the histogram methods are
indeed performing the nontrivial task of exploring the liquid-gas
phase boundary of the square-well fluid. For the purposes of this
paper, we will neglect to demonstrate that this phase transition is
indeed between liquid and gas phases, and take this fact as given.

\begin{figure}[!b]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/internal-energy.pdf}
  \caption[Internal energy]
  {Specific internal energy curves averaged over 30 $N=25$
    simulations.}
  \label{fig:internal_energy}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/heat-capacity.pdf}
  \caption[Heat capacity]
  {Specific heat capacity curves averaged over 30 $N=25$ simulations.}
  \label{fig:heat_capacity}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/config-entropy.pdf}
  \caption[Configurational entropy]
  {Specific configurational entropy curves averaged over 30 $N=25$
    simulations.}
  \label{fig:config_entropy}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/u-err.pdf}
  \caption[Internal energy error scaling]
  {Errors in specific internal energy, averaged over 30 $N=25$
    simulations.}
  \label{fig:internal_energy_err}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/cv-err.pdf}
  \caption[Heat capacity error scaling]
  {Errors in specific heat capacity, averaged over 30 $N=25$
    simulations.}
  \label{fig:heat_capacity_err}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.75\textwidth]{figs/s-err.pdf}
  \caption[Configurational entropy error scaling]
  {Errors in specific configurational entropy, averaged over 30 $N=25$
    simulations.}
  \label{fig:config_entropy_err}
\end{figure}

Having a sense of how $U$, $C_V$ and $S_{\t{config}}$ behave with
respect to temperature, we can now look at the how the errors $\Delta
U$, $\Delta C_V$ and $\Delta S_{\t{config}}$ relative to the TMMC
golden calculations for each method (except Wang-Landau). Figures
\ref{fig:internal_energy_err}--\ref{fig:config_entropy_err} show these
errors. Much like the arrays in Figure \ref{fig:arrays}, we provide
these figures only for a sense of how error curves behave. Errors (at
$kT/\epsilon\ge0.2$) typically peak near the phase transition at
$kT/\epsilon\approx0.5$, and drop back down.

Figures \ref{fig:u_scaling}--\ref{fig:s_scaling} provide a more
comprehensive picture of errors from each method. These figures show
the average maximum error in $U$, $C_V$, and $S$ for different system
sizes. Though errors jump around considerably in these figures, the
fact that they are roughly level between $N=15$ to $25$ justifies the
choice $N^2$ scaling for our number of initialization iterations.
Future studies of larger systems might use this scaling for initial
estimates of initialization times.

\begin{figure}[p]
  \centering
  \includegraphics[width=0.7\textwidth]{figs/u-scaling.pdf}
  \caption[Internal energy error scaling]
  {Specific internal energy error scaling.}
  \label{fig:u_scaling}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/cv-scaling.pdf}
  \caption[Heat capacity error scaling]
  {Specific heat capacity error scaling.}
  \label{fig:cv_scaling}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.7\textwidth]{figs/s-scaling.pdf}
  \caption[Configurational entropy error scaling]
  {Specific configurational entropy error scaling.}
  \label{fig:s_scaling}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.75\textwidth]{figs/u-comps.pdf}
  \caption[Internal energy error comparisons]
  {Specific internal energy errors, plotted against those from
    converged flat histogram weight simulations.}
  \label{fig:u_comps}
\end{figure}

\begin{figure}[p]
  \centering
  \includegraphics[width=0.7\textwidth]{figs/cv-comps.pdf}
  \caption[Heat capacity error comparisons]
  {Specific heat capacity errors, plotted against those from converged
    flat histogram weight simulations.}
  \label{fig:cv_comps}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=0.7\textwidth]{figs/s-comps.pdf}
  \caption[Configurational entropy error comparisons]
  {Specific configurational entropy errors, plotted against those from
    converged flat histogram weight simulations.}
  \label{fig:s_comps}
\end{figure}

It is more meaningful to compare errors against errors, rather than
system size. \fixme{wording} Figures
\ref{fig:u_comps}--\ref{fig:s_comps} show the errors of each method
plotted against errors of the converged flat histogram weight
simulations. The fact that some points fall below the diagonal
reference line is strange, and indicates that some regular simulations
are outperforming the converged flat ones. We do not understand why
such events occur, as converged flat simulations should be using
high-quality golden data to construct weight arrays, and should
therefore outperform all standard simulations. These events are thus
worthwhile looking into in more detail.

Figures \ref{fig:u_scaling}--\ref{fig:s_comps} make clear, again, the
inability of Wang-Landau to sample a system without a predefined
energy range. These figures also show that the performances of simple
flat method, vanilla Wang Landau, and TMMC are about the same, except
for an anomalous case with $N=21$ in which simple flat yields
simulation results with large errors. We do not know the reason for
this anomalous case, which might be worth investigating. Finally,
these error figures further confirm earlier suspicions of OE-TMMC's
underperformance relative to TMMC, despite the fact that it is
supposed to provide an improvement over the TMMC weight array.

@@

While the results in this paper are not strongly conclusive, they may
be used as guidelines for future studies of histogram methods,
especially as applied to the square-well fluid. After implementing
several histogram methods and applying them to the $\lambda=1.3$,
$\eta=0.3$ square-well fluid, we found that Wang-Landau, together with
a predefined energy range, consistently yields the smallest errors and
highest low-energy sampling rates. Without prior knowledge of the
minimum important energy of a system, however, Wang-Landau can get
stuck at low energies, and is incapable of recovering. Wang-Landau is
therefore most appropriate for studying systems in which the entire
energy range is known a priori, or in which only a given energy range
is of particular interest.

The simple flat method and TMMC methods seem to perform roughly on par
with vanilla Wang-Landau (i.e. Wang-Landau with a specified energy
range), except an anomalous case of failure worth investigating on the
part of simple flat. Our implementation of the optimized ensemble
using a transition matrix, which is meant to improve the TMMC weight
array, seemed to only make the weight array worse.

\fixme{fix when I'm awake}
